Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-06-22 paper/paperv3.qmd
bayesian segmented regression works!

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10,
         age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old"))) %>% 
  group_by(animalID) %>% 
  mutate(longevity = ifelse(max(year) < 2020,
                            max(age),
                            NA),
         longevity10 = longevity / 10,
         recruit_age = ifelse(max(year) < 2020 && any(observed_int == 1),
                              min(age[observed_int == 1]),
                              NA)) %>% 
  ungroup()

@@

# Fit GLMMs to the data with binary variable indicating before and after onset # of senescence. We a priori chose age 11 for senescence (age_senesce).
breed_age_mod <- glmer(
  observed_int ~ age_cat + age10 : age_cat + (1 | animalID) + (1 | year_fct),
  sealdat,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
breed_pred_pop <- ggpredict(breed_age_mod, 
                            terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

# Create WEIGHTED-AVERAGE predictions using years' random effects
seal_years <- count(sealdat, year_fct)
breed_pred_wgt <- ggpredict(
  breed_age_mod,
  terms = c("age10 [all]", "age_cat", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = group,
         year_fct = facet) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20) %>% 
  left_join(seal_years, by = "year_fct") %>% 
  group_by(age, age_cat) %>% 
  summarize(predicted = weighted.mean(predicted, n),
            .groups = "drop")

@@

ggsave("outputs/breeding_success.svg", height = 6, width = 10, units = "in")
```

@@

# Fit pup *survival* model
pup_surv_mod <- glmer(
  pup_surv_int ~ age10 * age_cat + (1 | animalID) + (1 | year_fct),
  surv_recr_data,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

# Fit pup *recruitment* model (female offspring only)
pup_recr_mod <- glmer(
  # NOTE: no random effect for individual bc singular fit
  # In survival model, 93 individuals (15%) had >3 pups (male+female pups)
  # In recruit model, only 17 individuals (4%) had >3 pups (female pups only)
  pup_recr_int ~ age10 * age_cat + (1 | year_fct),
  filter(surv_recr_data, pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

# Create pup survival POPULATION-LEVEL (i.e. without rand effects) predictions
surv_pred_pop <- ggpredict(
  pup_surv_mod, 
  terms = c("age10 [all]", "age_cat"),
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = group) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

@@

# Create pup recruit POPULATION-LEVEL (i.e. without rand effects) predictions
recr_pred_pop <- ggpredict(
  pup_recr_mod, 
  terms = c("age10 [all]", "age_cat"),
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = group) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

@@

@@

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
sexrat_pred <- ggpredict(sexrat_age_mod, 
                         terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce))

>>>>

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = (age - age_senesce) / 10,
         age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old"))) %>% 
  group_by(animalID) %>% 
  mutate(longevity = ifelse(max(year) < 2020,
                            max(age),
                            NA),
         longevity10 = longevity / 10,
         recruit_age = ifelse(max(year) < 2020 && any(observed_int == 1),
                              min(age[observed_int == 1]),
                              NA)) %>% 
  ungroup()

@@

# Fit GLMMs to the data with binary variable indicating before and after onset # of senescence. We a priori chose age 11 for senescence (age_senesce).
breed_age_mod <- glmer(
  observed_int ~ age10 : age_cat + (1 | animalID) + (1 | year_fct),
  sealdat,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
breed_pred_pop <- ggpredict(breed_age_mod, 
                            terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

# Create WEIGHTED-AVERAGE predictions using years' random effects
seal_years <- count(sealdat, year_fct)
breed_pred_wgt <- ggpredict(
  breed_age_mod,
  terms = c("age10 [all]", "age_cat", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = group,
         year_fct = facet) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20) %>% 
  left_join(seal_years, by = "year_fct") %>% 
  group_by(age, age_cat) %>% 
  summarize(predicted = weighted.mean(predicted, n),
            .groups = "drop")

@@

@@

# Fit pup *survival* model
pup_surv_mod <- glmer(
  pup_surv_int ~ age10 : age_cat + (1 | animalID) + (1 | year_fct),
  surv_recr_data,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

# Fit pup *recruitment* model (female offspring only)
pup_recr_mod <- glmer(
  # NOTE: no random effect for individual bc singular fit
  # In survival model, 93 individuals (15%) had >3 pups (male+female pups)
  # In recruit model, only 17 individuals (4%) had >3 pups (female pups only)
  pup_recr_int ~ age10 : age_cat + (1 | year_fct),
  filter(surv_recr_data, pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

# Create pup survival POPULATION-LEVEL (i.e. without rand effects) predictions
surv_pred_pop <- ggpredict(
  pup_surv_mod, 
  terms = c("age10 [all]", "age_cat"),
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = group) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

@@

# Create pup recruit POPULATION-LEVEL (i.e. without rand effects) predictions
recr_pred_pop <- ggpredict(
  pup_recr_mod, 
  terms = c("age10 [all]", "age_cat"),
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = group) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

@@

sexrat_age_mod <- glmer(
  is_male ~ age10 : age_cat + (1 | animalID) + (1 | year_fct),
  pup_sex_data,
  family = "binomial"
)

@@

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
sexrat_pred <- ggpredict(sexrat_age_mod, 
                         terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce))
